Binary companions of evolved stars in \apogee\ \DR: \\ Search method and catalog of $\sim$5,000 companions

### 2018-04-24 paper/1-catalog/twoface-catalog.tex
colon capitalization

We have solved the problem of deriving comprehensive multi-modal \pdf\ samplings
previously with \thejoker\ (\citealt{Price-Whelan:2017}), a Monte Carlo
rejection sampler that is computationally expensive but probabilistically
righteous:
It delivers independent posterior \pdf\ samples for single-companion binary
orbital parameters, given any number of radial-velocity measurements.
Here we use \thejoker\ to generate posterior \pdf\ samples for stars observed by
the \apogee\ survey (see \sectionname~\ref{sec:data}; \citealt{Majewski:2017}).

@@

Binary and multiple star systems are of great interest in astrophysics: The
population of stars and their companions encodes information about star
formation processes, stellar parameters and evolution, and the dynamics of
multi-body systems (for recent reviews, see \citealt{Duchene:2013,Moe:2017}).
Most of what is known about stellar companions comes from studies of nearby
main-sequence (MS) stars (e.g.,
\citealt{Duquennoy:1991,Raghavan:2010,Tokovinin:2014,Moe:2017}).
Nearly 50\% of MS stars in the solar neighborhood have companions
(e.g., \citealt{Tokovinin:2014}).
MS stars with companions have a large dynamic range of constituent and orbital
characteristics.
For example, binary stars have mass-ratios that span from $\approx 0.03$
to 1 (e.g., \citealt{Kraus:2008}), and have periods from days to
millions of years (e.g., \citealt{Raghavan:2010}).
Less is known about population properties of non-interacting or detached
companions to evolved stars.
The catalogs and methodology presented in this work are a first step towards
performing population inferences of binary star systems with evolved star
members.

@@

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in Keplerian orbital parameters, given radial-velocity
measurements under the assumptions listed above.
Our parametrization of the orbital elements is similar to the notation in
\citet{Murray:2010}:
The radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, M_0, t_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,M_0,\omega,K,v_0)$---period, eccentricity, mean
anomaly at a reference time $t_0$, argument of pericenter, velocity
semi-amplitude, barycentric velocity---and the true anomaly, $f$, is a function
of time and the specified parameters (see \sectionname~2 of
\citealt{Price-Whelan:2017} or \eqname~63 in \citealt{Murray:2010}).
In addition to the orbital parameters listed above, \thejoker\ can also generate
samples in an ``excess variance'' parameter, $s^2$, that is added to the
per-visit measurement variances.
This parameter allows us to test whether the visit RV uncertainties are
underestimated
For upper \RGB\ stars the inferred excess variance will be a combination of
extra systematic uncertainty and true astrophysical surface jitter
(e.g., \cite{Hekker:2008}).

\thejoker\ was designed for the extremely multi-modal \pdf 's expected when the
number of radial-velocity measurements of a source is small, or the data are
sparse (in phase-coverage) or noisy.
While other Markov Chain Monte Carlo (MCMC) methods have difficulty producing
independent samples with such data, \thejoker\ succeeds by brute force:
After generating an initial (very large) library of prior samples from an
assumed prior \pdf\ (see below), the (typically multi-modal) likelihood is
evaluated at each sample and used to reject all prior samples above the
likelihood surface (the rejection sampling step).
In practice, given a number of requested samples for each star, the sampling
proceeds iteratively: since it is easier to accept samples when the data is
sparse or noisy, far more prior sample draws (and thus likelihood evaluations)
must occur under very constraining data.

@@

We request \nposterior\ posterior samples for each source.
Depending on the data quality and phase coverage of the visits, \thejoker\ will
require different numbers of prior samples in order to rejection-sample down to
the requested number of posterior samples: For few-epoch or noisy RV data, many
prior samples will pass the rejection step, whereas for very precise or
many-epoch RV data, \thejoker\ may need to process the full library of prior
samples.
We therefore generate the posterior samples using an iterative procedure that
adaptively predicts how many prior samples to test for each star.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples (as few as one sample).
When this occurs, we continue sampling either using standard MCMC, or by
increasing the size of the prior cache and continuing rejection sampling with
\thejoker.

@@

% \todo{Rix suggests this needs 1-2 more sentences to ease the transition: "It comes as an unexpected abrupt turn in the logical flow."}
With posterior samplings for all \apogee\ \DR\ systems in hand, and as an
initial use of the per-source posterior samplings, we use a hierarchical
Bayesian model to infer the parameters of the (assumed Gaussian) prior over the
log-excess-variance parameter, $(\mu_y, \sigma_y)$ (see
\tablename~\ref{tbl:params}).
This inference serves as a test-case for future work, where we intend to use the
independent posterior samplings to construct a hierarchical inference over
companion population properties.
This is also a test of the visit velocity uncertainties reported in the \apogee\
data products: If the catalog uncertainties are significantly underestimated, we
expect the inferred log-excess-variance distribution parameters to tend towards
larger values.

@@

However, there are a number of caveats to keep in mind about this estimate of
the excess variance distribution.
First, we do not remove triple or other multiple systems: For any individual
system, the radial velocity variations induced by other massive bodies will lead
to larger preferred values for the excess variance parameter.
We do not expect there to be a large number of triple systems in our sample, but
this is still an important consideration for future efforts.
Second, we are sensitive to outliers and very non-Gaussian systematic error
distributions.
We later assess the non-Gaussianity of the visit velocity uncertainties by
looking at the visit velocity residuals away from the orbit samples produced by
\thejoker\ using the updated excess variance distribution (see
\sectionname~\ref{sec:discuss-assumptions}).

@@

We do not expect a sharp transition in inferred orbital parameters between stars
with and without companions: There is a continuum of companion properties.
For example, the companions can be low mass, or at long periods, or at high
inclination, all of which will make the velocity semi-amplitude, $K$, small for
a given system.
Therefore, there is no simple cut on radial velocity data alone that would
select a complete sample of stars with companions.
It is nevertheless possible to define a cut that selects stars with
high-confidence companions.

@@

The period-eccentricity plot (upper left panel in
\figurename~\ref{fig:logg-teff-p-e}) shows a clear signature of tidal
circularization: Such plots for main sequence stars typically show a more
distinct circularization period, below which the majority of stars have
eccentricities close to zero.
Here, we see that close to $P\approx 10~\textrm{day}$ the majority of companion
orbits appear to be circular, but between 10--100~days there appears to be a
gradual decrease in eccentricity rather than a sharp transition.
This is likely because the primary stars in this sample have a large range in
surface gravities and therefore a large range in stellar radii (see lower right
panel).

@@

In \figurename~\ref{fig:mass}, a few interesting systems are immediately
obvious: From the left panel, two systems have bimodal period samplings in which
both period modes put the minimum companion mass below the Hydrogen-burning
limit (brown dwarf candidates), and two systems have $M_{2, \textrm{min}} > M_1$
(neutron star or black hole candidates).
In the right panel, one system appears right on the limit of being an
interacting binary ($q_{\textrm{min}} \approx 0.5$), but has no significant UV
flux (\acronym{GALEX} DR5; \citealt{Bianchi:2011}).
These systems are prime candidates for spectroscopic follow-up.

@@

Paradoxically, there appear to exist short period ($\lesssim 100~\textrm{day}$)
companions at low surface gravities ($\logg \lesssim 1$), obvious in the
lower-left panel of \figurename~\ref{fig:logg-teff-p-e}: These companions would
orbit within the surface of their primary stars.
Each of these systems appears fine from the \apogee\ data quality flags, but
could nonetheless have incorrect stellar parameters.
As a test, we would ideally be able to compare the spectroscopic stellar
parameters to an independent determination from, e.g., asteroseismology.
Unfortunately, none of these short-period, low-\logg\ stars in the unimodal or
bimodal samples appear in the \acronym{APOKASC} (\citealt{Serenelli:2017})
catalog, which provides asteroseismic stellar parameters for a few hundred
\apogee\ dwarf and giant stars.
We therefore instead cross-match all high-$K$ stars with $\logg < 1$ and $99\%$
of their period samples below $100~\textrm{day}$ to the \acronym{APOKASC}
catalog: One star appears in both samples 2M19024490+4419523.
In \apogee\ \DR, this star has $\logg = 0.57$, but asteroseismic $\logg > 4$
(e.g., \texttt{HUBER\_LOGG}), consistent with being an M dwarf.
Many of these systems could therefore be low-mass dwarf stars with incorrect
stellar parameters in \apogee\ \DR.

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{c | c}
    \multicolumn{2}{c}{\textbf{Parent sample of \apogee\ \DR\ stars}}\\
    \hline
    \texttt{APOGEE\_ID} & \texttt{lnK\_per\_1} \\
    \hline
    2M00000002+7417074 & -2.028 \\
    2M00000068+5710233 & -2.600 \\
    2M00000222+5625359 &  2.134 \\
    2M00000446+5854329 & -7.199 \\
    ... & ... \\
    \hline
    \multicolumn{2}{c}{\textit{(\nstars\ rows)}}
    \end{tabular}
    \caption{This table contains \texttt{APOGEE\_ID}'s for all \nstars\ stars
    in the parent sample used in this work, for which we have posterior
    samplings in orbital parameters.
    The other column contains the 1st percentile computed over the $\ln K$
    samples for each source, \texttt{lnK\_per\_1}:
    The ``high-$K$'' sample (see \sectionname~\ref{sec:catalog-confident}) is
    defined using this column, by selecting \texttt{lnK\_per\_1 > 0}.
    }
    \label{tbl:lnK-per}
\end{table}

>>>>

We have solved the problem of deriving comprehensive multi-modal \pdf\ samplings
previously with \thejoker\ (\citealt{Price-Whelan:2017}), a Monte Carlo
rejection sampler that is computationally expensive but probabilistically
righteous:
it delivers independent posterior \pdf\ samples for single-companion binary
orbital parameters, given any number of radial-velocity measurements.
Here we use \thejoker\ to generate posterior \pdf\ samples for stars observed by
the \apogee\ survey (see \sectionname~\ref{sec:data}; \citealt{Majewski:2017}).

@@

Binary and multiple star systems are of great interest in astrophysics: the
population of stars and their companions encodes information about star
formation processes, stellar parameters and evolution, and the dynamics of
multi-body systems (for recent reviews, see \citealt{Duchene:2013,Moe:2017}).
Most of what is known about stellar companions comes from studies of nearby
main-sequence (MS) stars (e.g.,
\citealt{Duquennoy:1991,Raghavan:2010,Tokovinin:2014,Moe:2017}).
Nearly 50\% of MS stars in the solar neighborhood have companions
(e.g., \citealt{Tokovinin:2014}).
MS stars with companions have a large dynamic range of constituent and orbital
characteristics.
For example, binary stars have mass-ratios that span from $\approx 0.03$
to 1 (e.g., \citealt{Kraus:2008}), and have periods from days to
millions of years (e.g., \citealt{Raghavan:2010}).
Less is known about population properties of non-interacting or detached
companions to evolved stars.
The catalogs and methodology presented in this work are a first step towards
performing population inferences of binary star systems with evolved star
members.

@@

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in Keplerian orbital parameters, given radial-velocity
measurements under the assumptions listed above.
Our parametrization of the orbital elements is similar to the notation in
\citet{Murray:2010}:
the radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, M_0, t_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,M_0,\omega,K,v_0)$---period, eccentricity, mean
anomaly at a reference time $t_0$, argument of pericenter, velocity
semi-amplitude, barycentric velocity---and the true anomaly, $f$, is a function
of time and the specified parameters (see \sectionname~2 of
\citealt{Price-Whelan:2017} or \eqname~63 in \citealt{Murray:2010}).
In addition to the orbital parameters listed above, \thejoker\ can also generate
samples in an ``excess variance'' parameter, $s^2$, that is added to the
per-visit measurement variances.
This parameter allows us to test whether the visit RV uncertainties are
underestimated
For upper \RGB\ stars the inferred excess variance will be a combination of
extra systematic uncertainty and true astrophysical surface jitter
(e.g., \cite{Hekker:2008}).

\thejoker\ was designed for the extremely multi-modal \pdf 's expected when the
number of radial-velocity measurements of a source is small, or the data are
sparse (in phase-coverage) or noisy.
While other Markov Chain Monte Carlo (MCMC) methods have difficulty producing
independent samples with such data, \thejoker\ succeeds by brute force:
after generating an initial (very large) library of prior samples from an
assumed prior \pdf\ (see below), the (typically multi-modal) likelihood is
evaluated at each sample and used to reject all prior samples above the
likelihood surface (the rejection sampling step).
In practice, given a number of requested samples for each star, the sampling
proceeds iteratively: since it is easier to accept samples when the data is
sparse or noisy, far more prior sample draws (and thus likelihood evaluations)
must occur under very constraining data.

@@

We request \nposterior\ posterior samples for each source.
Depending on the data quality and phase coverage of the visits, \thejoker\ will
require different numbers of prior samples in order to rejection-sample down to
the requested number of posterior samples: for few-epoch or noisy RV data, many
prior samples will pass the rejection step, whereas for very precise or
many-epoch RV data, \thejoker\ may need to process the full library of prior
samples.
We therefore generate the posterior samples using an iterative procedure that
adaptively predicts how many prior samples to test for each star.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples (as few as one sample).
When this occurs, we continue sampling either using standard MCMC, or by
increasing the size of the prior cache and continuing rejection sampling with
\thejoker.

@@

% \todo{Rix suggests this needs 1-2 more sentences to ease the transition: "It comes as an unexpected abrupt turn in the logical flow."}
With posterior samplings for all \apogee\ \DR\ systems in hand, and as an
initial use of the per-source posterior samplings, we use a hierarchical
Bayesian model to infer the parameters of the (assumed Gaussian) prior over the
log-excess-variance parameter, $(\mu_y, \sigma_y)$ (see
\tablename~\ref{tbl:params}).
This inference serves as a test-case for future work, where we intend to use the
independent posterior samplings to construct a hierarchical inference over
companion population properties.
This is also a test of the visit velocity uncertainties reported in the \apogee\
data products: if the catalog uncertainties are significantly underestimated, we
expect the inferred log-excess-variance distribution parameters to tend towards
larger values.

@@

However, there are a number of caveats to keep in mind about this estimate of
the excess variance distribution.
First, we do not remove triple or other multiple systems: for any individual
system, the radial velocity variations induced by other massive bodies will lead
to larger preferred values for the excess variance parameter.
We do not expect there to be a large number of triple systems in our sample, but
this is still an important consideration for future efforts.
Second, we are sensitive to outliers and very non-Gaussian systematic error
distributions.
We later assess the non-Gaussianity of the visit velocity uncertainties by
looking at the visit velocity residuals away from the orbit samples produced by
\thejoker\ using the updated excess variance distribution (see
\sectionname~\ref{sec:discuss-assumptions}).

@@

We do not expect a sharp transition in inferred orbital parameters between stars
with and without companions: there is a continuum of companion properties.
For example, the companions can be low mass, or at long periods, or at high
inclination, all of which will make the velocity semi-amplitude, $K$, small for
a given system.
Therefore, there is no simple cut on radial velocity data alone that would
select a complete sample of stars with companions.
It is nevertheless possible to define a cut that selects stars with
high-confidence companions.

@@

The period-eccentricity plot (upper left panel in
\figurename~\ref{fig:logg-teff-p-e}) shows a clear signature of tidal
circularization: such plots for main sequence stars typically show a more
distinct circularization period, below which the majority of stars have
eccentricities close to zero.
Here, we see that close to $P\approx 10~\textrm{day}$ the majority of companion
orbits appear to be circular, but between 10--100~days there appears to be a
gradual decrease in eccentricity rather than a sharp transition.
This is likely because the primary stars in this sample have a large range in
surface gravities and therefore a large range in stellar radii (see lower right
panel).

@@

In \figurename~\ref{fig:mass}, a few interesting systems are immediately
obvious: from the left panel, two systems have bimodal period samplings in which
both period modes put the minimum companion mass below the Hydrogen-burning
limit (brown dwarf candidates), and two systems have $M_{2, \textrm{min}} > M_1$
(neutron star or black hole candidates).
In the right panel, one system appears right on the limit of being an
interacting binary ($q_{\textrm{min}} \approx 0.5$), but has no significant UV
flux (\acronym{GALEX} DR5; \citealt{Bianchi:2011}).
These systems are prime candidates for spectroscopic follow-up.

@@

Paradoxically, there appear to exist short period ($\lesssim 100~\textrm{day}$)
companions at low surface gravities ($\logg \lesssim 1$), obvious in the
lower-left panel of \figurename~\ref{fig:logg-teff-p-e}: these companions would
orbit within the surface of their primary stars.
Each of these systems appears fine from the \apogee\ data quality flags, but
could nonetheless have incorrect stellar parameters.
As a test, we would ideally be able to compare the spectroscopic stellar
parameters to an independent determination from, e.g., asteroseismology.
Unfortunately, none of these short-period, low-\logg\ stars in the unimodal or
bimodal samples appear in the \acronym{APOKASC} (\citealt{Serenelli:2017})
catalog, which provides asteroseismic stellar parameters for a few hundred
\apogee\ dwarf and giant stars.
We therefore instead cross-match all high-$K$ stars with $\logg < 1$ and $99\%$
of their period samples below $100~\textrm{day}$ to the \acronym{APOKASC}
catalog: one star appears in both samples 2M19024490+4419523.
In \apogee\ \DR, this star has $\logg = 0.57$, but asteroseismic $\logg > 4$
(e.g., \texttt{HUBER\_LOGG}), consistent with being an M dwarf.
Many of these systems could therefore be low-mass dwarf stars with incorrect
stellar parameters in \apogee\ \DR.

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{c | c}
    \multicolumn{2}{c}{\textbf{Parent sample of \apogee\ \DR\ stars}}\\
    \hline
    \texttt{APOGEE\_ID} & \texttt{lnK\_per\_1} \\
    \hline
    2M00000002+7417074 & -2.028 \\
    2M00000068+5710233 & -2.600 \\
    2M00000222+5625359 &  2.134 \\
    2M00000446+5854329 & -7.199 \\
    ... & ... \\
    \hline
    \multicolumn{2}{c}{\textit{(\nstars\ rows)}}
    \end{tabular}
    \caption{This table contains \texttt{APOGEE\_ID}'s for all \nstars\ stars
    in the parent sample used in this work, for which we have posterior
    samplings in orbital parameters.
    The other column contains the 1st percentile computed over the $\ln K$
    samples for each source, \texttt{lnK\_per\_1}:
    the ``high-$K$'' sample (see \sectionname~\ref{sec:catalog-confident}) is
    defined using this column, by selecting \texttt{lnK\_per\_1 > 0}.
    }
    \label{tbl:lnK-per}
\end{table}
